---
title: "GapMinder Exercises"
author: "Markus Loecher, Berlin School of Economics and Law"
output:
   html_document:
    #variant: markdown_github
    toc: true
    number_sections: true
    self_contained: no
    toc_depth: 2
    toc_float: true
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# library(gapminder)
# library(dplyr)
# library(ggplot2)
library(pacman)
p_load(gapminder)
p_load(dplyr)
p_load(ggplot2)
p_load(magrittr) # JUST in case it did not load from the previous calls

p_load(googleVis)
#p_load(ISLR)
EVAL = FALSE
```

# Data Manipulation 

* Add a `filter()` line after the pipe (`%>%`) to extract only the observations from the year 1957.  Remember that you use `==` to compare two values.

```{r, eval=EVAL}
# Filter the gapminder dataset for the year 1957
gapminder %>%
```

* Filter the `gapminder` data to retrieve only the observation from China in the year 2002.

```{r}
# Filter for China in 2002
```


* Sort the `gapminder` dataset in ascending order of life expectancy (`lifeExp`).
* Sort the `gapminder` dataset in descending order of life expectancy.

```{r}
# Sort in ascending order of lifeExp


  
# Sort in descending order of lifeExp
```

* Use `filter()` to extract observations from just the year 1957, then use `arrange()` to sort in descending order of population (`pop`).

```{r}
# Filter for the year 1957, then arrange in descending order of population
```


* Use `mutate()` to change the existing `lifeExp` column, by multiplying it by 12: `12 * lifeExp`.
* Use `mutate()` to add a new column, called `lifeExpMonths`, calculated as `12 * lifeExp`.


```{r}
# Use mutate to change lifeExp to be in months


# Use mutate to create a new column called lifeExpMonths
```


In one sequence of pipes on the `gapminder` dataset:
* `filter()` for observations from the year 2007,
* `mutate()` to create a column `lifeExpMonths`, calculated as `12 * lifeExp`, and
* `arrange()` in descending order of that new column

```{r}
# Filter, mutate, and arrange the gapminder dataset
```

* Filter `gapminder` for observations from the year 1952, and assign it to a new dataset `gapminder_1952` using the assignment operator (`<-`).

```{r}
# Create gapminder_1952
```

# Data Visualization


* Change the scatter plot of `gapminder_1952` so that (`pop`) is on the x-axis and GDP per capita (`gdpPercap`) is on the y-axis.


```{r, eval= EVAL}
# Change to put pop on the x-axis and gdpPercap on the y-axis
ggplot(gapminder_1952, aes(x = gdpPercap, y = lifeExp)) +
  geom_point()
```

* Create a scatter plot of `gapminder_1952` with population (`pop`) is on the x-axis and life expectancy (`lifeExp`) on the y-axis.

```{r}
# Create a scatter plot with pop on the x-axis and lifeExp on the y-axis

```

* Change the existing scatter plot (code provided) to put the x-axis (representing population) on a log scale.

```{r, eval= EVAL}
# Change this plot to put the x-axis on a log scale
ggplot(gapminder_1952, aes(x = pop, y = lifeExp)) +
  geom_point()
```

* Create a scatter plot with population (`pop`) on the x-axis and GDP per capita (`gdpPercap`) on the y-axis. Put both the x- and y- axes on a log scale.

```{r}
# Scatter plot comparing pop and gdpPercap, with both axes on a log scale
```

* Create a scatter plot with population (`pop`) on the x-axis, life expectancy (`lifeExp`) on the y-axis, and with continent (`continent`) represented by the color of the points. Put the x-axis on a log scale.

```{r}
# Scatter plot comparing pop and lifeExp, with color representing continent
```

* Modify the scatter plot so that the size of the points represents each country`s GDP per capita (`gdpPercap`).

```{r}
# Add color representing continent
# Add the size aesthetic to represent a country`s gdpPercap

```

## Facets

Create a scatter plot of gapminder_1952 with the x-axis representing population (`pop`), the y-axis representing life expectancy (`lifeExp`), and faceted to have one subplot per continent (`continent`). Put the x-axis on a log scale.

```{r}
# Scatter plot comparing pop and lifeExp, faceted by continent

```


* Put GDP per capita (`gdpPercap`) on the x-axis and life expectancy (`lifeExp`) on the y-axis, with continent (`continent`) represented by color and population (`pop`) represented by size.

* Put the x-axis on a log scale 

* Facet by the `year` variable



```{r}
# Scatter plot comparing gdpPercap and lifeExp, with color representing continent
# and size representing population, faceted by year

```

# Grouping and summarizing

* Use the `median()` function within a `summarize()` to find the median life expectancy. Save it into a column called `medianLifeExp`


```{r}
# Summarize to find the median life expectancy
```

* Filter for the year 1957, then use the `median()` function within a `summarize()` to calculate the median life expectancy into a column called `medianLifeExp`

* Find the median life expectancy (`lifeExp`) and maximum GDP per capita (`gdpPercap`) within each year, saving them into `medianLifeExp` and `maxGdpPercap`, respectively.

* Filter the `gapminder` data for the year 1957. Then find the median life expectancy (`lifeExp`) and maximum GDP per capita (`gdpPercap`) **within each continent**, saving them into `medianLifeExp` and `maxGdpPercap`, respectively.

```{r}
#
```


* Find the median life expectancy (`lifeExp`) and maximum GDP per capita (`gdpPercap`) within each combination of continent and year, saving them into `medianLifeExp` and `maxGdpPercap`, respectively.

```{r}
# Find median life expectancy and maximum GDP per capita in each continent/year combination
```

# June 15th, 2020

* Use the `by_year` dataset to create a scatter plot showing the change of median life expectancy over time, with year on the x-axis and `medianLifeExp` on the y-axis. Be sure to add `expand_limits(y = 0)` to make sure the plot's y-axis includes zero.

```{r}
# Create a scatter plot showing the change in medianLifeExp over time

```



* Summarize the `gapminder` dataset by continent and year, finding the median GDP per capita (`gdpPercap`) within each and putting it into a column called `medianGdpPercap`. Use the assignment operator `<-` to save this summarized data as `by_year_continent`.

* Create a scatter plot showing the change in `medianGdpPercap` by continent over time. Use color to distinguish between continents, and be sure to add `expand_limits(y = 0)` so that the y-axis starts at zero.

```{r}
# Summarize medianGdpPercap within each continent within each year: by_year_continent

# Plot the change in medianGdpPercap in each continent over time

```


* Filter the gapminder dataset for the year 2007, then summarize the median GDP per capita and the median life expectancy **within each continent*, into columns called `medianLifeExp` and `medianGdpPercap.` Save this as `by_continent_2007.`

* Use the `by_continent_2007` data to create a scatterplot comparing these summary statistics for continents in 2007, putting the median GDP per capita on the x-axis to the median life expectancy on the y-axis. Color the scatter plot by `continent.` You don`t need to add `expand_limits(y = 0)` for this plot.


```{r}
# Summarize the median GDP and median life expectancy per continent in 2007


# Use a scatter plot to compare the median GDP and median life expectancy

```

## The famous "Hans Rosling chart" 

https://www.ted.com/talks/hans_rosling_the_best_stats_you_ve_ever_seen#t-241405

```{r}
M1 <- gvisMotionChart(Fruits, idvar="Fruit", timevar="Year")
plot(M1)
```

* Modify the command above to replicate 

> You've never seen data presented like this. With the drama and urgency of a sportscaster, statistics guru Hans Rosling debunks myths about the so-called "developing world."
