---
title: "Simple Modeling in R"
author: "Markus Loecher, Berlin School of Economics and Law"
output:
   html_document:
    #variant: markdown_github
    toc: true
    number_sections: true
    self_contained: yes
    toc_depth: 2
    toc_float: true
    fig_caption: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
baseR = FALSE
library(pander)
library(dplyr)
library(tidyverse)
library(nycflights13)
```

```{r}
#define our own transformation
require(scales) # trans_new() is in the scales library
sign_sqrt_trans = function() trans_new("sign_sqrt", function(x) sign(x)*sqrt(abs(x)), function(x) sign(x)*x^2)

```

# Simple data summaries

## Titanic


```{r, echo = TRUE }
train <- read.csv("data/TitanicTrain.csv")
```


The disaster was famous for saving "women and children first", so let's take a look at the Sex and Age variables to see if any patterns are evident. We'll start with the gender of the passengers. After reloading the data into R, take a look at the summary of this variable:

```{r, echo = TRUE }
round(prop.table(table(train$Sex, train$Survived),1),2)
```


Let's create a new variable, "Child", to indicate whether the passenger is below the age of 18:

```{r, echo = TRUE }
train$Child <- 0
train$Child[train$Age < 18] <- 1
```

Now we want to create a table with both gender and age to see the survival proportions for different subsets.
Recall the dplyr version of the aggregate function:

```{r, echo = TRUE}

#summarise(group_by(train, Sex, Child), round(mean(Survived),2), length(Survived))
options(digits=3)

train %>% group_by(Sex, Child) %>%
  summarize(propSurvived = mean(Survived),
            sampleSize = length(Survived))

```


While the class variable is limited to a manageable 3 values, the fare is again a continuous variable that needs to be reduced to something that can be easily tabulated. Let's bin the fares into less than $10, between $10 and $20, $20 to $30 and more than $30 and store it to a new variable:

```{r, echo = TRUE}
train$Fare2 <- '30+'
 train$Fare2[train$Fare < 30 & train$Fare >= 20] <- '20-30'
 train$Fare2[train$Fare < 20 & train$Fare >= 10] <- '10-20'
train$Fare2[train$Fare < 10] <- '<10'
```

* Use the summarise function to compute the proportions and  in each group defined by child, gender and Fare2:

```{r, echo = FALSE}

options(digits=3)

SurvProp <- train %>% group_by(Child, Fare2, Sex) %>%
  summarize(propSurvived = mean(Survived),
            sampleSize = length(Survived)) %>% arrange(propSurvived)

SurvProp
```

* Find strong differences between male/female survival probabilities.
* Partition your data into even finer subgroups, e.g. age and/or embarkation port and repeat.

```{r, echo = FALSE}
#summarise(group_by(train, Child, Fare2, Sex, Embarked), round(mean(Survived),2), length(Survived))

options(digits=3)

train %>% group_by(Child, Fare2, Sex, Embarked) %>%
  summarize(propSurvived = mean(Survived),
            sampleSize = length(Survived))

```

# Linear Models

```{r, echo = TRUE}
fit = lm(Survived ~ Sex, data = train)
summary(fit)
```

```{r, echo = TRUE}
fit = lm(Survived ~ Sex + Pclass, data = train)
summary(fit)
```
Try to reframe our exercises from above into linear models.

* Think about 
  * factors
  * significance
  

## Correlation 

#### Are arrival and departure delay correlated?

```{r, fig.width=8}
set.seed(Sys.Date())
ranRows=sample(nrow(flights),5000)

if (baseR){
  plot(arr_delay ~ dep_delay, data=flights[ranRows,], pch=20,cex=0.5,col=carrier)
  grid()
} else {
  p = ggplot(flights[ranRows,], aes( dep_delay,arr_delay, col=carrier ))
  p + geom_point( alpha=0.5, size=1) + coord_trans(x="sign_sqrt", y="sign_sqrt") # +  geom_smooth(method=lm) 
}


cor(flights$dep_delay,flights$arr_delay, use = "complete.obs",method = "pearson")
#slooow    
#cor(flights$dep_delay,flights$arr_delay, use = "complete.obs",method = "kendall")
  
#cor(flights$dep_delay,flights$arr_delay, use = "complete.obs",method = "spearman")

Shortdelays = filter(flights, abs(arr_delay)<100 & abs(dep_delay)<100)

cor(Shortdelays$dep_delay,Shortdelays$arr_delay, use = "complete.obs",method = "pearson")

#plot(arr_delay ~ dep_delay,data=Shortdelays)
```


## simple regression 

### mpg vs. weight


```{r}
library(ISLR);data(Auto)
plot(mpg ~ weight, data = Auto,col = rgb(0,0,1,0.5), pch=20,xlim=c(250, 7000), ylim = c(0,45));grid()
LSfit = lm(mpg ~ weight, data = Auto)

#overlay regression line
abline(LSfit, col=2)

#summary
summary(LSfit)

#diagnostics
#plot(LSfit, c(1,2,5))

```

**Tasks**

1. Predict mpg for a car that weighs 5000 lbs.
2. Confidence Interval for slope
3. Are the least squares assumptions met?
4. Add 3 outliers to the data: 
  * weight=500, mpg = 40
  * weight=$10^4$, mpg = 80
  * weight=3000, mpg = 80

Points that fall horizontally far from the line are points of high leverage; these points
can strongly influence the slope of the least squares line. If one of these high leverage
points does appear to actually invoke its in
uence on the slope of the line then we call it an in
uential point. Usually we can say
a point is influential if, had we fitted the line without it, the influential point would have
been unusually far from the least squares line.


```{r, fig.width=12, fig.height = 6}
par(mfrow=c(1,3), cex=1.4)

outlrs = cbind(weight=c(500,10000,3000),mpg=c(40,80,80))
Auto2 = Auto

fit=list()
for (i in 1:3){
  Auto2[1,c("weight","mpg")] = outlrs[i,]
  
  plot(mpg ~ weight, data = Auto2,col = rgb(1,0.894,0.769,0.5), pch=20);grid()
  #overlay regression line
  abline(LSfit, col=2, lwd=2.5)
  fit[[i]] = lm(mpg ~ weight, data = Auto2) 
  points(Auto2[1,c("weight","mpg")], col = i+2, pch = 18, cex = 2)
  #overlay regression line
  abline(fit[[i]], col=i+2,lwd=2, lty=2)
  #plot(fit[[i]],5)
}


```

  